Tail associations and their impact on extinction risk

### 2019-10-29 MainText.Rmd
more edits

<!--ECEs have impacts in the spatial context as well-->
ECEs also have impacts in ecological scenarios for which spatial dynamics are
important.
In the metapopulation context, spatial synchrony of population dynamics, which can
be caused by spatial synchrony of climatic fluctuations, is linked to metapopulation
extinction risk [@hanski1998metapopulation; @earn2000coherence].
If ECEs are also spatially more extensive than moderate events,
as seems likely for some types of events, extreme events may accentuate
population synchrony and thereby increase extinction risk.
For instance, a 21-year-long
study of @tack2015increasing on the Glanville fritillary butterfly demonstrated that frequency 
of drought during early larval instars increased the strength and extent 
of spatial synchrony 
over a metapopulation network of 4000 dry meadows, and could thereby influence
the long-term viability of the species in the area of study.

@@

<!--ECEs probably produce tail associations-->
There are reasons to believe that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. For instance, if a single ECE is associated
with spatially widespread, large values 
of some climatic variable, but small values of the
variable are associated with local weather rather than ECEs, the result could
be stronger right-tail association between measurements of the climatic variable
made in different locations, through time.
For example, rainstorms which deposit exceptionally large amounts of rain
may do so over large areas; but months with less rain may experience
scattered showers only, which may be much less correlated across locales.

@@

There are also reasons to hypothesize that asymmetry of tail association in 
environmental variables may influence extinction risks for metapopulations.
For instance, assuming for the sake of a simple example that low values of an environmental
variable are "bad" for populations of a given species and high values are "good",
greater left-tail associations between measurements of the variable in different
habitat patches may cause higher metapopulation extinction risk because then
very bad years for component populations would occur at the same time in 
many patches, reducing the potential for rescue effects. Greater right-tail associations
would probably not have the same effect, even if overall correlations between locations
were the same, because in that case it would be very good years for the component 
populations that would occur at the same time in many patches [@Ghosh_copula]. 
@Ghosh_copula substantiated this intuition by showing that it held true for a 
modeling setup using a spatial version of a very simple non-density-dependent
population growth model (the Lewontin-Cohen model). In essence, if environmental
"catastrophes" (i.e., extremely bad years for a population) are widely 
spatially synchronized it should create much greater metapopulation extinction 
risk than if "bonanzas" (i.e., extremely good years) are widely synchronized, even
if overall (non-tail-specific) levels of environmental synchrony are the same
in both these cases. The present study extends tests of this idea to metapopulation
models incorporating density dependence.

Extinction risk is also very commonly estimated for age-structured models 
such as stochastic matrix models [@MorrisDoak2002; @Caswell2006] and 
integral projection models
[@Merowetal2014]. For such models, extinction risk depends, in part, on the long-term
population growth rate, a population parameter which in turn 
depends on the means, variances and covariances of population vital rates such 
as the survival and fecundity rates of age classes. Extinction risks may,
*a priori*, also depend on possible asymmetries of tail association between
population vital rates. For instance, it seems reasonable to imagine
a case in which both adult fecundity and juvenile survival rates 
depend on an environmental variable such as spring rainfall in similar ways, but 
only when that variable takes very low values. When rainfall takes
typical or high values, fecundity and survival may be limited instead 
by different factors. This could produce left-tail dependence between these 
vital rates, possibly resulting in different extinction risk for the model
compared to a case in which vital rates were symmetrically associated in 
their tails but otherwise statistically similar.

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
instead of attempting mathematical analysis and/or using models that are more realistic, 
because our intention here is to provide an initial exploration only of 
the potential mechanisms outlined above which suggest that
asymmetries of tail association may make a difference for extinction risks. 
Some material in the Discussion outlines possible next steps.
<!--***DAN: In Discussion need to talk about what analytic results are needed
to take the next step of determining when tail depencence may make enough of
a difference to think about. I think our results, together with Tulja's 
approximation, suggest that tail associations probably don't make much difference
for models without a spatial component. But our results illsutaret that it can
make a big difference when there is a spatial component, apparently because of 
the influence of tail associations on rescue effects. Could maybe even outline
analytic approaches for non-dd models? Not sure what would be done with dd
models. The point is you need to take the next step in the Discussion of 
talking about how our results, which admitedly are just for specific models,
could be turned into analytic results or general results. Interestingly, 
Tulja's approximation may also apply to the spatial case, at least when 
all locations can be reached from all other locations via dispersal. But the 
approximation seems poor in that case. Need to think about it more, and maybe
need more research.-->
Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are associated with
tail dependence in environmental variables.

@@

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two stochastic quantities can have same Kendall 
correlation, but different degrees of tail association: 
(A) extreme left-tail association, (B) moderate left-tail association, (C) 
moderate right-tail association, and (D) extreme right-tail association.
In all cases, Kendall's $\tau$ was the same, $\tau = 0.75$,
up to sample variation (the sample correlation for each dataset is displayed
on its panel). The extreme left-tail (respectively, right-tail) dependent 
case is perfectly correlated below (respectively, above) the threshold of 0.
See Methods for information on how bivariate noise was generated with
asymmetric tail associations.
\label{fig_intro_cop}}
\end{center}
\end{figure*}
<!--***DAN: Shyamolina, it occured to be there are simple and good analytic reasons
why ktau_common is exactly 0.75 (up to sample variation). 
I bet you will see them too if you look up the
definition of kendall tau and think about the extreme tail dependent cases. 
Currently, around line 50 of the sup mat Rmd, you calculate ktau_common
by simulating 100000 points from the extreme-tail-denepdnece noise and then
calculating the sample kendall. Instead, can you please replace those lines 
with the simple line "ktau_common<-0.75"? I would do this myself but I
don't want to risk messing up your code in some way I don't understand. You should not
need the seed anymore to ensure you get the same ktau_common on each run
since you won't be using the sample correlartion anymore to estimate ktau_common,
you'll just be using the analytic result that the population correlation
of data from the extreme-tail-dependent algorithm (in the limit of infinite
sample sizes) is exactly 0.75. Please
let me know if you do not understand something here.-->
<!--***DAN: Shyamolina, currently the sample kendalls are the same on A and D, and
also on B and C, and I know why. This looks weird, and makes the reader
wonder why this is the case. Rather than plotting the reversed version of the
data from A on D, and the reversed version of the data from B on C, can you plase
generate separate, independent data for A and D, and also for B and C? That
will have the effect of making the sample kendalls on the 4 panels slightly
different, which will mean readers wont think anything funny is going on.-->

@@

The stochastic quantities $s_e(t)$ and $\tilde{f}_a(t)$ were generated 
to have the same overall association (Kendall's $\tau$ equal to $0.75$,
up to sampling variation), but different degrees and types of 
asymmetry of tail association in different simulations, as follows.
For a simulation of length $T$, we used methods that can 
generate $T$ independent pairs $(u(t),v(t))$ for $t=1,\ldots,T$ with
$\tau=0.75$ (up to sample variation), but with either extreme left-,
moderate left-, moderate right-, or extreme right-tail association 
(e.g., Fig. \ref{fig_intro_cop}). The methods we used generated
$(u(t),v(t))$ such that $u(t)$ was uniformly distributed on $(0,1)$,
and likewise for $v(t)$. We then applied a monotonic transformation
to the $u(t)$ so that the resulting numbers were distributed according to
a beta distribution with parameters
$\alpha=3$ and $\beta=2$ 
and used those values for the $s_e(t)$. We also applied a (different)
monotonic transformation to the $v(t)$ so that the resulting numbers were
distributed according to a gamma distribution with parameters $k=2$
and $\theta=1$ and used those values for the $\tilde{f}_a(t)$. Because
the $u(t)$ and $v(t)$ were uniformly distributed, the inverse 
cummulative distribution functions of the desired beta and
gamma distributions were the transformations. Because these
were monotonic transformations, the Kendall correlation of 
$s_e(t)$ and $\tilde{f}_a(t)$ was the same as that of 
$u(t)$ and $v(t)$. The quantities $(u(t),v(t))$ for $t=1,\ldots,T$ were
generated as described in Appendix \ref{SM-makenoise}.

@@

<!--***DAN: Insert para here talking about in what sense the variables
$s_e(t)$ and $f_a(t)$ are asymmetrically associated, through time, in their 
tails. This is where you intro the partial Spearman and may also make references
to Genest and others indicating that, when you are interested in dependence,
you want a rank-based notion of association - it's the only thing that
separates the association from the marginals. I will insert some text now,
but it won't fit perfectly until after we go through and replace Kendall with
Spearman throughout.-->
The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$ and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$ (smallest elements
are defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which meet with the unit square
(Fig. \ref{PartialSpearmanFig}). 
<!--***DAN: need to make this figure-->
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. 
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the ranks, $x$ and $y$, lie between the boundary lines
we defined (Fig. \ref{PartialSpearmanFig}) in
the unit square. The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by a monotonic transformation (and likewise for 
$v(t)$ and $\tilde{f}_a(t)$). Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed.

@@

<!--density-dependence metapopulation model-->
To investigate Q2 from the \nameref{Introduction}, we considered spatialized, stochastic versions
of six density dependent population models, the Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, 
and Austin-Brewer models (@cohen1995unexpected; Table \ref{tab_model_summary}).
All models took the general form $\vec{P} (t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{\text{th}}$ component $P_i(t)$ representing
the population density in the $i^{\text{th}}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{\text{th}}$ diagonal entry 
$\lambda_i(t)$ depnding on the model used (see Table \ref{tab_model_summary}). The matrix $D$
was an $n \times n$ dispersal matrix. For simplicity, 
we assumed the $n$ habitat patches were arranged, evenly spaced, in a line.
We considered both "local" and "global" dispersal among patches in different simulations. 
For local dispersal, a fraction $d$ of 
each population dispersed during each time step, equally distributed to
the two or one nearest neighbors. For global dispersal the same fraction $d$ dispersed
equally to the other $(n-1)$ patches. When $d$ was $0$, dispersal did not occur and 
rescue effects were not possible.
Two models that @cohen1995unexpected used were excluded from our analyses because they were not
readily adapted to our simulation setup: the Varley model,
which is not continuous;
<!--***DAN: Shyamolina, looking back at the Cohen paper, using the value C=1 as they used, the
Varley model *is* continuous. It is not differentiable, but it's not clear to me that 
non-differentiability means we cannot use the model. Perhaps it does mean that, but I don't see 
it. Let's discuss. I'm not saying we have to go back and use it, but we may have to justify
differently why we used 6 of Cohen's models and not the other two.-->
and the Malthus-Condorcet-Mill model, which employs a logarithm of the population and is therefore
inconvenient to analyze in the context of extinction.
<!--***DAN: Unless I am mistaken, the limit of the right hand side of the M-C-M model as
described by Cohen, as P approaches 0 from the right, is 0. That means the model has
a removable singularity at 0, i.e., we could just define the model so P_t+1=0 when P_t
is 0, and we would get a perfectly valid continuous model. This means that using logs is 
not really a valid reason to not use the models. Let's discuss. I'm not saying we have to go 
back and use it, but we may have to justify differently why we used 6 of Cohen's models and not 
the other two.-->

@@

<!--***DAN: Dan is starting to wonder if referees will wonder something like "what 
happens to tail associations when you do these transformatons." We do not wonder 
this because we have in mind a defn for tail associations that only depends on ranks,
but we have not presented that defn. We have instead only given an intuitive description
of tail association. This may be a problem. Referees will see uf generating data with tail associatons
and uniform marginals, but then transforming it, and putting the transformed values into 
our models. And we are confortable with the idea that that transformation does not
alter the tail associations, but they are not comfortable with that.
We may need to define tail associations
using partial Spearman or something, to get this idea across. The problem with that is, 
we use Kendall throughout
the rest of the ms, and so if we define partial Spearman that will create an inconsistency
(using Kendall in some places, Spearman in others) that may raise eyebrows in a short paper 
like this. Not sure right now what to do anout this, we need to think about it. Can we add a new
para at the beginnign of methods giving some kind of formal introduction to tail associations
that makes clear it is something we consider to be for ranks? Without using partial Spearman?
There is no partial Kendall, since Kendall is based on concordant and discordant pairs.
An alterntive is to replace Kendall wth Spearman throughout the paper, then go ahead
and intro partial Spearmn as a way of making it clear tail associaton is for ranks. 
Is this possible? It would definitely be a pain, so therefore want to avoid unless
necessary. One thing this would require would be calculating the populaton (not sample) 
Spearman of the extreme tail associated data. At least we could estimate it very precisely
using a huge sample (would not have the same problems doing that as we did for Kendall),
and may even be able to calculate it analytically. Let's just use Spearman and calculate it very
precisely using a large sample. Well, maybe we could calculate it analytically. 
-->

@@

<!--***DAN: Dan stopped here on 2019 10 27 until we resolve some of the issues that have come up-->

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of six density dependence metapopulation models considered for each patch; last three columns indicate equilibrium population value ($P_e^{i}$), critical growth rate ($r_c^{i}$ : transition point from under- to over-compensatory dynamics of deterministic models) and bifurcation parameter ($r_{bf}^{i}$ : transition point from stable to unstable equilibrium of deterministic models), respectively. Superscript $i=1,2,...,n$ indicates index for $i^{th}$ patch.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```
<!--***DAN: Shyamolina, in this table and in the sup mat (and anywhere else where this comment applies),
please replace the notation P^i_t and \epsilon^i_t by P_i(t) and \epsilon_i(t), the reason being,
putting an i in the superscript makes it look like and exponent. This is legitimately confusing because
we use lots of other exponents. I want you to generally refrain from placing the abitat patch index in the
superscript, putting in the subscript instead. This means you have to put the t in parens in front of the
variable in question, instead of in the subscript. This will require several changes to this table. Please
make those changes and then commit. Then please make the corresponding changes to the sup mat, whereever needed.-->
<!--***DAN: Dan, once Shyamolina does the above, you will then have to go back and edit this table in various
ways that you cannot do until after that change is made.-->

>>>>

<!--ECEs have impacts in the spatial context as well-->
ECEs also have impacts in ecological contexts for which spatial dynamics are
important.
In the metapopulation context, spatial synchrony of population dynamics, which can
be caused by spatial synchrony of climatic fluctuations [@Liebhold2004], is linked to metapopulation
extinction risk [@hanski1998metapopulation; @earn2000coherence].
If ECEs are also spatially more extensive than moderate events,
as seems likely for some types of events, extreme events may accentuate
population synchrony and thereby increase extinction risk.
For instance, a 21-year-long
study of @tack2015increasing on the Glanville fritillary butterfly demonstrated that frequency 
of drought during early larval instars increased the strength and extent 
of spatial synchrony 
over a metapopulation network of 4000 dry meadows, and could thereby influence
the long-term viability of the species in the area of study.

@@

<!--***DAN: maybe you want to have a Discussion para on copulas. It could start with
some text about ranks being the proper way to study associations, and then could go
on to link to copulas and cite BIVAN-->

<!--ECEs probably produce tail associations-->
There are reasons to hypothesize that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. For instance, if a single ECE is associated
with spatially widespread, large values 
of some climatic variable, but small values of the
variable are associated with local weather rather than ECEs, the result could
be stronger right-tail association, through time, between measurements of the climatic variable
made in different locations.
For example, rainstorms which deposit exceptionally large amounts of rain
may do so over large areas; but periods with less rain may experience
scattered showers only, which may be much less correlated across locales.

@@

There are also reasons to hypothesize that asymmetry of tail association in 
environmental variables may influence extinction risks for metapopulations.
For instance, assuming for the sake of a simple example that low values of an environmental
variable are "bad" for populations of a given species and high values are "good",
greater left-tail associations between measurements of the variable in different
habitat patches may cause higher metapopulation extinction risk because then
very bad years for component populations would occur at the same time in 
many patches, reducing the potential for rescue effects. Greater right-tail associations
would probably not have the same effect, even if overall correlations between locations
were the same, because in that case it would be very good years for the component 
populations that would occur at the same time in many patches. 
@Ghosh_copula substantiated this intuition by showing that it held true for a 
modeling setup using a spatial version of a very simple non-density-dependent
population growth model (the Lewontin-Cohen model). In essence, if environmental
"catastrophes" (i.e., extremely bad years for a population) are widely 
spatially synchronized it should create much greater metapopulation extinction 
risk than if "bonanzas" (i.e., extremely good years) are widely synchronized, even
if overall (non-tail-specific) levels of environmental synchrony are the same
in both these scenarios. The present study extends tests of this idea to metapopulation
models incorporating density dependence.

Extinction risk is also very commonly estimated for age-structured models 
such as stochastic matrix models [@MorrisDoak2002; @Caswell2006] and 
integral projection models
[@Merowetal2014]. For such models, extinction risk depends, in part, on the long-term
population growth rate, a population parameter which in turn 
depends on the means, variances and covariances of population vital rates such 
as the survival and fecundity rates of age classes. Extinction risks may,
*a priori*, also depend on possible asymmetries of tail association between
population vital rates. For instance, it seems reasonable to imagine
a case in which both adult fecundity and juvenile survival rates 
depend on an environmental variable such as spring rainfall in similar ways, but 
only when that variable takes low values. When rainfall takes
typical or high values, fecundity and survival may be limited instead 
by different factors. This could produce left-tail dependence between these 
vital rates, possibly resulting in different extinction risk for the model
compared to a case in which vital rates were symmetrically associated in 
their tails but otherwise statistically similar.

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
instead of attempting mathematical analysis and/or using models that are more realistic, 
because our intention here is to provide an initial exploration only of 
the potential mechanisms, outlined above, which suggest that
asymmetries of tail association may make a difference for extinction risks. 
By so doing we hope to open these potentially practically important questions to 
wider examination other researchers. Some material in the Discussion outlines some possible next steps.
<!--***DAN: In Discussion need to talk about what analytic results are needed
to take the next step of determining when tail depencence may make enough of
a difference to think about. I think our results, together with Tulja's 
approximation, suggest that tail associations probably don't make much difference
for models without a spatial component. But our results illsutaret that it can
make a big difference when there is a spatial component, apparently because of 
the influence of tail associations on rescue effects. Could maybe even outline
analytic approaches for non-dd models? Not sure what would be done with dd
models. The point is you need to take the next step in the Discussion of 
talking about how our results, which admitedly are just for specific models,
could be turned into analytic results or general results. Interestingly, 
Tulja's approximation may also apply to the spatial case, at least when 
all locations can be reached from all other locations via dispersal. But the 
approximation seems poor in that case. Need to think about it more, and maybe
need more research.-->
Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are related to
tail associations in environmental variables.

@@

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two stochastic quantities can have same Spearman 
correlation, but different degrees of tail association: 
(A) extreme left-tail association, (B) moderate left-tail association, (C) 
moderate right-tail association, and (D) extreme right-tail association.
In all cases, Spearman's $\rho$ was the same, $\rho = 0.875$,
up to sample variation (the sample correlation for each dataset is displayed
on its panel). The extreme left-tail (respectively, right-tail) dependent 
case is perfectly correlated below (respectively, above) the threshold of 0.
See Methods for information on how bivariate noise was generated with
asymmetric tail associations.
\label{fig_intro_cop}}
\end{center}
\end{figure*}

@@

The stochastic quantities $s_e(t)$ and $\tilde{f}_a(t)$ were generated 
to have the same overall association (Spearman's $\rho$ equal to $0.875$,
up to sampling variation), but different degrees and types of 
asymmetry of tail association in different simulations, as follows.
For a simulation of length $T$, we used methods that can 
generate $T$ independent pairs $(u(t),v(t))$ for $t=1,\ldots,T$ with
$\rho=0.875$ (up to sample variation), but with either extreme left-,
moderate left-, moderate right-, or extreme right-tail association 
(e.g., Fig. \ref{fig_intro_cop}). The methods we used generated
$(u(t),v(t))$ such that $u(t)$ was uniformly distributed on $(0,1)$,
and likewise for $v(t)$. We then applied a monotonic transformation
to the $u(t)$ so that the resulting numbers were distributed according to
a beta distribution with parameters
$\alpha=3$ and $\beta=2$, 
and we used those values for the $s_e(t)$. We also applied a (different)
monotonic transformation to the $v(t)$ so that the resulting numbers were
distributed according to a gamma distribution with parameters $k=2$
and $\theta=1$, and we used those values for the $\tilde{f}_a(t)$. Because
the $u(t)$ and $v(t)$ were uniformly distributed, the inverse 
cummulative distribution functions of the desired beta and
gamma distributions were the transformations. Because these
were monotonic transformations, the Spearman correlation of 
$s_e(t)$ and $\tilde{f}_a(t)$ was the same as that of 
$u(t)$ and $v(t)$. The quantities $(u(t),v(t))$ for $t=1,\ldots,T$ were
generated as described in Appendix \ref{SM-makenoise}.

@@

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$ and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which meet with the unit square
(Fig. \ref{PartialSpearmanFig}). 
<!--***DAN: need to make this figure-->
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. 
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation (and likewise for 
$v(t)$ and $\tilde{f}_a(t)$). Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables.

@@

<!--density-dependence metapopulation model-->
To investigate Q2 from the \nameref{Introduction}, we considered spatialized, stochastic versions
of six density dependent population models, the Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, 
and Austin-Brewer models (Table \ref{tab_model_summary}).
All models took the general form $\vec{P} (t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{\text{th}}$ component $P_i(t)$ representing
the population density in the $i^{\text{th}}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{\text{th}}$ diagonal entry 
$\lambda_i(t)$ depending on the model used (Table \ref{tab_model_summary}). The matrix $D$
was an $n \times n$ dispersal matrix. For simplicity, 
we assumed the $n$ habitat patches were arranged, evenly spaced, in a line.
We considered both "local" and "global" dispersal among patches in different simulations. 
For local dispersal, a fraction $d$ of 
each population dispersed during each time step, equally distributed to
the two or one nearest neighbors. For global dispersal the same fraction $d$ dispersed
equally to the other $n-1$ patches. When $d$ was $0$, dispersal did not occur and 
rescue effects were not possible.

<!--***DAN: Also need to talk about how you selected parameters-->
We obtained our list of classic population models (which we then spatialized and made 
stochastic) from @cohen1995unexpected. 
Model parameters were selected so that varying $r$ across a range, while keeping
the other selected parameters fixed at selected values for a model, led to a 
transition from undercompensatory to overcompensatory equilibrium dynamics
for the deterministic, one-patch versions of the models. Whereas @cohen1995unexpected
was interested in studying chaotic model behavior and therefore chose parameters
for which the models exhibited chaotic dynamics, we were interested in non-chaotic model
behaviors, which seem likely to be more common in real populations. 
Two models that @cohen1995unexpected 
used were excluded from our analyses because their functional forms complicated 
the analyses we wished to perform. First, the equilibrium of the 
Malthus-Condorcet-Mill model [@cohen1995unexpected] could not be determined 
analytically, making downstream analysis difficult. Second, for
the Varley model and for the portions of parameter space we explored, 
varying $r$ produced no transition from under- to overcompensatory dynamics.

<!--***DAN: got to here-->

@@

@@

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of six density dependence metapopulation models considered for each patch; last three columns indicate equilibrium population value ($P_e^{i}$), critical growth rate ($r_c^{i}$ : transition point from under- to over-compensatory dynamics of deterministic models) and bifurcation parameter ($r_{bf}^{i}$ : transition point from stable to unstable equilibrium of deterministic models), respectively. Superscript $i=1,2,...,n$ indicates index for $i^{th}$ patch.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```
<!--***DAN: Shyamolina, in this table and in the sup mat (and anywhere else where this comment applies),
please replace the notation P^i_t and \epsilon^i_t by P_i(t) and \epsilon_i(t), the reason being,
putting an i in the superscript makes it look like and exponent. This is legitimately confusing because
we use lots of other exponents. I want you to generally refrain from placing the habitat patch index in the
superscript, putting in the subscript instead. This means you have to put the t in parens in front of the
variable in question, instead of in the subscript. This will require several changes to this table. Please
make those changes and then commit. Then please make the corresponding changes to the sup mat, whereever needed.-->
<!--***DAN: Dan, once Shyamolina does the above, you will then have to go back and edit this table in various
ways that you cannot do until after that change is made.-->
